---
title: "Day 1: Laying the groundwork"
date: "Point count data analysis workshop 2025"
author: "Peter Solymos"
fontsize: 11pt
urlcolor: blue
output:
  beamer_presentation:
    theme: "Montpellier"
    # colortheme: "dolphin"
    # fonttheme: "structurebold"
    # theme: "Singapore"
    slide_level: 2
    incremental: false
    includes:
      in_header: ../preamble.tex
---

```{r setup,include=FALSE}
options(width=53, scipen=999)
library(knitr)
```

# Introduction and background

## Intro

- Camera on optional
- Go for a break as needed (2 breaks/session)
- Stretch


## Questions:

- Live
- In the chat
- Or https://hackmd.io/@psolymos/qpad-2021

**Juan Andres** (TA) will organize these, thanks!

***

## About the instructor

- **Peter Solymos**
- Stat ecologist,
- R programmer (vegan, detect, ResourceSelection, pbapply),
- Sometimes I teach
- [\@psolymos](https://twitter.com/psolymos)

***

## Acknowledgement

I am situated on Indigenous land in Treaty 6 territory;
land occupied, traveled, and cared for by Indigenous Peoples
since time immemorial to the present day.

***

## What you will learn

1. How to analyze your point count data when it combines different methodologies/protocols/technologies
2. How to violate assumptions and get away with it

***

## Outline

Day 1

- Introduction
- We need to talk about data
- A primer in regression techniques

Day 2

- Behavioral complexities

Day 3

- The detection process
- Dealing with recordings

Day 4

- Putting it all together
- Roadside surveys

***

## Install packages

```{r eval=FALSE}
source(
  paste0("https://raw.githubusercontent.com/",
  "psolymos/qpad-workshop/main/src/install.R")
)
```

***

## Get course materials

1. Visit [https://github.com/psolymos/qpad-workshop/releases](https://github.com/psolymos/qpad-workshop/releases)
2. Download the latest release into a folder
3. Extract the zip/tar.gz archive
4. Open the `workshop.Rproj` file in RStudio (or open any other R GUI/console and `setwd()` to the directory where you downloaded the file)
5. (You can delete the archive)

***

# Review of field sampling techniques

## Random sampling

xxx

## Straified

zzz
